% 11-roadmap.tex

\section{Research Roadmap}\label{sec:roadmap}

Phase~1 (foundation) is complete.
Subsequent phases produce independently evaluable artifacts.

\textbf{Phase~2: F$\star$X Core.}
Z3 proof suites for all core invariants, consteval infrastructure, reflection checks, proof-carrying build.
Eval: proof coverage, solver time scaling, structural check false positive rate.

\textbf{Phase~3: Longitude.}
GPU profiling, network probing, kernel calibration, Z3 topology solver.
Eval: calibration accuracy (predicted vs measured per shape/device), solver time at various cluster sizes, configuration quality vs expert tuning.

\textbf{Phase~4: Augur.}
Digital twin, drift detection, bottleneck diagnosis, model intelligence.
Eval: prediction accuracy, regression detection latency, diagnostic accuracy.

\textbf{Phase~5: Compiled Tiers~2--3.}
\code{ConductorTensorImpl}, async kernel execution, CUDA Graph replay.
Eval: per-op dispatch overhead (Tier~2 vs 1 vs eager), per-iteration overhead (Tier~3 vs 2), end-to-end throughput on standard benchmarks.

\textbf{Phase~6: Distribution.}
Canopy (gossip, Raft, CRDTs), Keepers, full Cipher (three-tier persistence, configurable $\alpha$ redundancy).
Eval: recovery time, redundancy overhead, heterogeneous scaling efficiency.

\textbf{Phase~7: Model-Aware Optimization.}
Token adaptation, layer analysis, architecture mutation, training optimization.
Integrated pipeline: Augur diagnoses, F$\star$X verifies, Keeper activates.
Eval: time-to-accuracy improvement, quality preservation under automated mutation, comparison with expert manual optimization.

Each phase produces a focused evaluation against established baselines.
We plan to make prototype, formalization, and evaluation artifacts publicly available.
